---
title: "Biostat 203B Homework 1"
author: Hua Zhou
subtitle: Due Jan 22 @ 11:59PM
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Display machine information for reproducibility:
```{r, eval=F}
sessionInfo()
```

## Q1. Git/GitHub

**No handwritten homework reports are accepted for this course.** We work with Git and GitHub. Efficient and abundant use of Git, e.g., frequent and well-documented commits, is an important criterion for grading your homework.

1. Apply for the [Student Developer Pack](https://education.github.com/pack) at GitHub using your UCLA email.

**Answer**: This step has been attempted, but the application has been rejected. The corresponding email can be supplied as desired.

2. Create a **private** repository `biostat-203b-2021-winter` and add `Hua-Zhou`, `Chris-German` and `ElvisCuiHan` as your collaborators with write permission.

**Answer**: The step has been completed and may be verified by the instructor and teaching assistants.

3. Top directories of the repository should be `hw1`, `hw2`, ... Maintain two branches `master` and `develop`. The `develop` branch will be your main playground, the place where you develop solution (code) to homework problems and write up report. The `master` branch will be your presentation area. Submit your homework files (R markdown file `Rmd`, `html` file converted from R markdown, all code and data sets to reproduce results) in `master` branch.

**Answer**: The step has been completed and may be verified by the instructor and teaching assistants.

4. After each homework due date, teaching assistant and instructor will check out your master branch for grading. Tag each of your homework submissions with tag names `hw1`, `hw2`, ... Tagging time will be used as your submission time. That means if you tag your `hw1` submission after deadline, penalty points will be deducted for late submission.

**Answer**: The step has been completed and may be verified by the instructor and teaching assistants.

5. After this course, you can make this repository public and use it to demonstrate your skill sets on job market.

**Answer**: I will attempt to do so when the time comes, provided I remember.

## Q2. Linux Shell Commands

1. This exercise (and later in this course) uses the [MIMIC-IV data](https://mimic-iv.mit.edu), a freely accessible critical care database developed by the MIT Lab for Computational Physiology. Follow the instructions at <https://mimic-iv.mit.edu/docs/access/> to (1) complete the CITI `Data or Specimens Only Research` course and (2) obtain the PhysioNet credential for using the MIMIC-IV data. Display the verification links to your completion report and completion certificate here. (Hint: The CITI training takes a couple hours and the PhysioNet credentialing takes a couple days; do not leave it to the last minute.)

**Completion Report**: citiprogram.org/verify/?k6c9b2a64-f8c9-43ae-b0be-d8b81e9ca8df-40373113
**Completion Certificate**: https://www.citiprogram.org/verify/?wf756585a-2dab-4399-99cf-695874c85a96-40373113

2. The `/usr/203b-data/mimic-iv/` folder on teaching server contains data sets from MIMIC-IV. Refer to <https://mimic-iv.mit.edu/docs/datasets/> for details of data files.  
    ```{bash,eval=F}
    ls -l /usr/203b-data/mimic-iv
    ```
Please, do **not** put these data files into Git; they are big. Do **not** copy them into your directory. Do **not** decompress the gz data files. These create unnecessary big files on storage and are not big data friendly practices. Just read from the data folder `/usr/203b-data/mimic-iv` directly in following exercises. 

    Use Bash commands to answer following questions.

3. Display the contents in the folders `core`, `hosp`, `icu`. What are the functionalities of the bash commands `zcat`, `zless`, `zmore`, and `zgrep`? 

**Answer**: The contents in the folders of the MIMIC IV dataset are listed. They appear to be compressed CSV's of the data; the biggest files occupy file sizes on the order of magnitude of 10-100 MB (it is likely that the files become much bigger when unzipped):

```{bash,eval=F}
ls /usr/203b-data/mimic4/core
```
    -rw-r--r--. 1 huazhou huazhou 17224843 Jan 11 01:55 admissions.csv.gz
    -rw-r--r--. 1 huazhou huazhou  2884996 Jan 11 01:54 patients.csv.gz
    -rw-r--r--. 1 huazhou huazhou 51188147 Jan 11 01:55 transfers.csv.gz

```{bash,eval=F}
ls /usr/203b-data/mimic4/hosp
```
    -rw-r--r--. 1 huazhou huazhou     430049 Jan 11 01:57 d_hcpcs.csv.gz
    -rw-r--r--. 1 huazhou huazhou   26575586 Jan 11 01:57 diagnoses_icd.csv.gz
    -rw-r--r--. 1 huazhou huazhou     723633 Jan 11 02:00 d_icd_diagnoses.csv.gz
    -rw-r--r--. 1 huazhou huazhou     564422 Jan 11 02:11 d_icd_procedures.csv.gz
    -rw-r--r--. 1 huazhou huazhou      14845 Jan 11 02:00 d_labitems.csv.gz
    -rw-r--r--. 1 huazhou huazhou   12913088 Jan 11 02:00 drgcodes.csv.gz
    -rw-r--r--. 1 huazhou huazhou  518077567 Jan 11 02:13 emar.csv.gz
    -rw-r--r--. 1 huazhou huazhou  479709397 Jan 11 01:59 emar_detail.csv.gz
    -rw-r--r--. 1 huazhou huazhou    1415469 Jan 11 02:11 hcpcsevents.csv.gz
    -rw-r--r--. 1 huazhou huazhou 2093725833 Jan 11 02:09 labevents.csv.gz
    -rw-r--r--. 1 huazhou huazhou   15896456 Jan 11 02:13 microbiologyevents.csv.gz
    -rw-r--r--. 1 huazhou huazhou  423170857 Jan 11 02:11 pharmacy.csv.gz
    -rw-r--r--. 1 huazhou huazhou  501822286 Jan 11 01:57 poe.csv.gz
    -rw-r--r--. 1 huazhou huazhou   23675550 Jan 11 01:57 poe_detail.csv.gz
    -rw-r--r--. 1 huazhou huazhou  367321152 Jan 11 02:00 prescriptions.csv.gz
    -rw-r--r--. 1 huazhou huazhou    4965027 Jan 11 02:11 procedures_icd.csv.gz
    -rw-r--r--. 1 huazhou huazhou    9579255 Jan 11 02:11 services.csv.gz


```{bash,eval=F}
ls /usr/203b-data/mimic4/icu
```
    -rw-r--r--. 1 huazhou huazhou 2264326210 Jan 11 02:25 chartevents.csv.gz
    -rw-r--r--. 1 huazhou huazhou   40440772 Jan 11 02:13 datetimeevents.csv.gz
    -rw-r--r--. 1 huazhou huazhou      56593 Jan 11 02:14 d_items.csv.gz
    -rw-r--r--. 1 huazhou huazhou    2628845 Jan 11 02:25 icustays.csv.gz
    -rw-r--r--. 1 huazhou huazhou  328835832 Jan 11 02:14 inputevents.csv.gz
    -rw-r--r--. 1 huazhou huazhou   35300863 Jan 11 02:25 outputevents.csv.gz
    -rw-r--r--. 1 huazhou huazhou   19362097 Jan 11 02:14 procedureevents.csv.gz


`zcat` unzips the relevant zipped file to standard output, and corresponds to `cat` for zipped files. `zless` is the `less` command on zipped files; analogously, `zmore` is the `more` command on zipped files, and `zgrep` is `grep` on zipped files. The commands were verified via running `man` on each of the commands, and testing them on some sample files in the MIMIC IV data.

4. What's the output of following bash script?
    ```{bash, eval=F}
    for datafile in /usr/203b-data/mimic-iv/core/*.gz
      do
        ls -l $datafile
      done
    ```
Display the number of lines in each data file using a similar loop.

**Answer**: The bash script lists the zipped files in long-format.

We obtain the number of lines in each data file with the following loop:
```{bash, eval=F}
for f in /usr/203b-data/mimic-iv/*/*.gz; do
    echo $f,$(zcat $f | wc -l)
done
```

This gives us the following output, with number of lines for each file, separated by comma:

    /usr/203b-data/mimic-iv/core/admissions.csv.gz,524521
    /usr/203b-data/mimic-iv/core/patients.csv.gz,383221
    /usr/203b-data/mimic-iv/core/transfers.csv.gz,2192964
    /usr/203b-data/mimic-iv/hosp/d_hcpcs.csv.gz,89201
    /usr/203b-data/mimic-iv/hosp/diagnoses_icd.csv.gz,4677925
    /usr/203b-data/mimic-iv/hosp/d_icd_diagnoses.csv.gz,86752
    /usr/203b-data/mimic-iv/hosp/d_icd_procedures.csv.gz,82764
    /usr/203b-data/mimic-iv/hosp/d_labitems.csv.gz,1626
    /usr/203b-data/mimic-iv/hosp/drgcodes.csv.gz,1168136
    /usr/203b-data/mimic-iv/hosp/emar.csv.gz,27590436
    /usr/203b-data/mimic-iv/hosp/emar_detail.csv.gz,56203138
    /usr/203b-data/mimic-iv/hosp/hcpcsevents.csv.gz,144859
    /usr/203b-data/mimic-iv/hosp/labevents.csv.gz,122289829
    /usr/203b-data/mimic-iv/hosp/microbiologyevents.csv.gz,1026114
    /usr/203b-data/mimic-iv/hosp/pharmacy.csv.gz,14747760
    /usr/203b-data/mimic-iv/hosp/poe.csv.gz,42526845
    /usr/203b-data/mimic-iv/hosp/poe_detail.csv.gz,3259645
    /usr/203b-data/mimic-iv/hosp/prescriptions.csv.gz,17021400
    /usr/203b-data/mimic-iv/hosp/procedures_icd.csv.gz,685415
    /usr/203b-data/mimic-iv/hosp/services.csv.gz,563707
    /usr/203b-data/mimic-iv/icu/chartevents.csv.gz,327363275
    /usr/203b-data/mimic-iv/icu/datetimeevents.csv.gz,6999317
    /usr/203b-data/mimic-iv/icu/d_items.csv.gz,3836
    /usr/203b-data/mimic-iv/icu/icustays.csv.gz,69620
    /usr/203b-data/mimic-iv/icu/inputevents.csv.gz,8869716
    /usr/203b-data/mimic-iv/icu/outputevents.csv.gz,4248829
    /usr/203b-data/mimic-iv/icu/procedureevents.csv.gz,689847


5. Display the first few lines of `admissions.csv.gz`. How many rows are in this data file? How many unique patients (identified by `subject_id`) are in this data file? What are the possible values taken by each of the variable `admission_type`, `admission_location`, `insurance`, `language`, `marital_status`, and `ethnicity`? Also report the count for each unique value of these variables. (Hint: combine Linux commands `zcat`, `head`/`tail`, `awk`, `uniq`, `wc`, and so on.)

**Answer**: we run `zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | head` to get the first few lines:

    subject_id,hadm_id,admittime,dischtime,deathtime,admission_type,admission_location,discharge_location,insurance,language,marital_status,ethnicity,edregtime,edouttime,hospital_expire_flag
    12427812,21593330,2184-01-06 11:51:00,2184-01-10 11:45:00,,URGENT,PHYSICIAN REFERRAL,HOME,Other,ENGLISH,,UNKNOWN,,,0
    14029832,22059088,2120-01-18 01:28:00,2120-01-20 16:13:00,,URGENT,TRANSFER FROM HOSPITAL,HOME,Other,ENGLISH,,OTHER,,,0
    14495017,22484010,2175-01-28 15:41:00,2175-01-29 16:00:00,,DIRECT EMER.,PHYSICIAN REFERRAL,HOME,Other,?,,WHITE,,,0
    13676048,23865469,2193-01-19 05:27:00,2193-01-24 18:59:00,,URGENT,PHYSICIAN REFERRAL,HOME,Other,?,MARRIED,WHITE,,,0
    13831972,27763544,2131-01-27 04:03:00,2131-01-27 05:39:00,,EU OBSERVATION,EMERGENCY ROOM,,Medicaid,ENGLISH,SINGLE,WHITE,2131-01-26 22:19:00,2131-01-27 05:39:00,0
    18523038,25414328,2142-08-26 17:14:00,2142-08-27 10:00:00,,DIRECT OBSERVATION,PROCEDURE SITE,,Other,ENGLISH,SINGLE,WHITE,,,0
    16705931,20580522,2174-10-24 11:30:00,2174-10-24 18:45:00,,DIRECT OBSERVATION,PHYSICIAN REFERRAL,,Other,ENGLISH,MARRIED,WHITE,,,0
    19963742,24951506,2171-07-04 15:58:00,2171-07-05 13:37:00,,AMBULATORY OBSERVATION,PACU,,Other,ENGLISH,SINGLE,UNKNOWN,,,0
    10903424,22568585,2181-01-31 13:09:00,2181-01-31 13:42:00,,EU OBSERVATION,EMERGENCY ROOM,,Other,ENGLISH,SINGLE,WHITE,,,0

Our output from Q4.4 tells us that admissions.csv.gz has 524520 rows, excluding the header. We obtain the number of unique patients with `zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | tail -n+2 | awk -F, '{print $1}' | sort -u | wc -l` to get 257366 unique patients.

We use `sort -u` predominantly to obtain our unique `admission_type`, `admission_location`, etc. (e.g. `zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | tail -n+2 | awk -F, '{print $6}' | sort | uniq -c` for `admission_type`, including counts).

The full code for obtaining this information follows:

```{bash,eval=F}
for vars in admission_type,6\
            admission_location,7\
            insurance,9\
            language,10\
            marital_status,11\
            ethnicity,12; do
    varname=`echo $vars | awk -F, '{print $1}'`
    varnum=`echo $vars | awk -F, '{print $2}'`
    echo Variable name: $varname
    zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | tail -n+2 | head | awk -v col="$varnum" -F, '{print $col}' | sort | uniq -c
    echo
done
```

This gives us the following output:

Variable name: admission_type

    7259 AMBULATORY OBSERVATION
    21595 DIRECT EMER.
    20014 DIRECT OBSERVATION
    72077 ELECTIVE
    100938 EU OBSERVATION
    158032 EW EMER.
    55532 OBSERVATION ADMIT
    41103 SURGICAL SAME DAY ADMISSION
    47970 URGENT

Variable name: admission_location

    60437
    191 AMBULATORY SURGERY TRANSFER
    10678 CLINIC REFERRAL
    246241 EMERGENCY ROOM
    379 INFORMATION NOT AVAILABLE
    4484 INTERNAL TRANSFER TO OR FROM PSYCH
    6075 PACU
    127648 PHYSICIAN REFERRAL
    8453 PROCEDURE SITE
    39150 TRANSFER FROM HOSPITAL
    4069 TRANSFER FROM SKILLED NURSING FACILITY
    16715 WALK-IN/SELF REFERRAL

Variable name: insurance

    50962 Medicaid
    171492 Medicare
    302066 Other

Variable name: language

    46231 ?
    478289 ENGLISH

Variable name: marital_status

    66133
    33762 DIVORCED
    195326 MARRIED
    180453 SINGLE
    48846 WIDOWED

Variable name: ethnicity

    1536 AMERICAN INDIAN/ALASKA NATIVE
    24522 ASIAN
    80526 BLACK/AFRICAN AMERICAN
    29887 HISPANIC/LATINO
    26844 OTHER
    3742 UNABLE TO OBTAIN
    19419 UNKNOWN
    338044 WHITE

## Q3. Who's popular in Price and Prejudice

1. You and your friend just have finished reading *Pride and Prejudice* by Jane Austen. Among the four main characters in the book, Elizabeth, Jane, Lydia, and Darcy, your friend thinks that Darcy was the most mentioned. You, however, are certain it was Elizabeth. Obtain the full text of the novel from <http://www.gutenberg.org/cache/epub/42671/pg42671.txt> and save to your local folder. 
    ```{bash, eval=FALSE}
    curl http://www.gutenberg.org/cache/epub/42671/pg42671.txt > pride_and_prejudice.txt
    ```
Do **not** put this text file `pride_and_prejudice.txt` in Git. Using a `for` loop, how would you tabulate the number of times each of the four characters is mentioned?

**Answer**: Tabulate by running a for loop for every word in the file, counting the number of times "Elizabeth", "Jane", "Lydia", and "Darcy" are mentioned for simplicity. Note that Elizabeth is referred to by other appelations, namely "Eliza" and "Lizzy" in the story, so we would need to count those as well. The limitation of this method is that it short counts the number of mentions by missing pronouns that refer to the characters, but plays it safe by not counting pronouns that may refer to other characters, or roundabout mentions of said characters (e.g. "her older sister" may refer to "Jane" or "Elizabeth", depending on the context). It also does not account for other appelations for Jane, Lydia, and Darcy (Darcy's first name "Fitzwilliam" is mentioned twice in the novel and is always accompanied by his last name; other mentions of "Fitzwilliam" refer to another character, Colonel Fitzwilliam). Furthermore, there may be a slight overcount for "Jane", as the author shares the same first name as the character.

The code used to obtain this information follows. Note that we have to generalize for punctuation and "'s" occurrences (e.g. "Elizabeth." and "Elizabeth's" should both count as a mention to Elizabeth); hence we include the Kleene star * to follow up on the requisite strings:

```{bash,eval=F}
#NOTE: Elizabeth also goes under the name "Lizzy" and "Eliza"
n_Liz=0
n_Jane=0
n_Lydia=0
n_Darcy=0

# for each line in pride and prejudice
while read line; do
    for w in $line; do
        if [[ ("$w" == "Elizabeth"*) || ("$w" == "Eliza"*) ||\
 ("$w" == "Lizzy"*) ]]; then
            n_Liz=$((n_Liz + 1))
        elif [[ "$w" == "Jane"* ]]; then
            n_Jane=$((n_Jane + 1))
        elif [[ "$w" == "Lydia"* ]]; then
            n_Lydia=$((n_Lydia + 1))
        elif [[ "$w" == "Darcy"* ]]; then
            n_Darcy=$((n_Darcy + 1))
        fi
    done
done < pride_and_prejudice.txt
echo "Elizabeth: $n_Liz"
echo "Jane: $n_Jane"
echo "Lydia: $n_Lydia"
echo "Darcy: $n_Darcy"
```

The results follow:

    Elizabeth: 730
    Jane: 293
    Lydia: 169
    Darcy: 417


Elizabeth is the most mentioned by a noticeable margin. Note that the counts differ slightly when compared with the results from `grep -c`, though the conclusions remain the same, in that Elizabeth leads in references:

```{bash,eval=F}
n_Liz=$(grep -c -e Eliza -e Lizzy pride_and_prejudice.txt)
n_Jane=$(grep -c Jane pride_and_prejudice.txt)
n_Lydia=$(grep -c Lydia pride_and_prejudice.txt)
n_Darcy=$(grep -c Darcy pride_and_prejudice.txt)
echo "Elizabeth: $n_Liz"
echo "Jane: $n_Jane"
echo "Lydia: $n_Lydia"
echo "Darcy: $n_Darcy"
```

    Elizabeth: 751
    Jane: 290
    Lydia: 166
    Darcy: 414


2. What's the difference between the following two commands?
    ```{bash eval=FALSE}
    echo 'hello, world' > test1.txt
    ```
    and
    ```{bash eval=FALSE}
    echo 'hello, world' >> test2.txt
    ```

**Answer**: The first command sends the string 'hello, world' to the file test1.txt. If test1.txt already exists, it will contain only the string 'hello, world'. The second command appends the string 'hello, world' to the file test2.txt. If test2.txt already exists, it will contain the text in it previously, plus 'hello, world' at the end of the last line in the file (note that appending does not contain a newline by default, which is something you have to include yourself if desired).

3. Using your favorite text editor (e.g., `vi`), type the following and save the file as `middle.sh`:
    ```{bash eval=FALSE}
    #!/bin/sh
    # Select lines from the middle of a file.
    # Usage: bash middle.sh filename end_line num_lines
    head -n "$2" "$1" | tail -n "$3"
    ```
Using `chmod` make the file executable by the owner, and run 
    ```{bash eval=FALSE}
    ./middle.sh pride_and_prejudice.txt 20 5
    ```
Explain the output. Explain the meaning of `"$1"`, `"$2"`, and `"$3"` in this shell script. Why do we need the first line of the shell script?

**Answer**: We use `chmod 771 middle.sh` to make the script readable and writable by anybody in the group, but not absolutely everybody, but allow the script to be executable by everybody. This is in case anybody outside the group rewrites the script to contain malicious code and then executes it, potentially erasing the whole server.

We then run `./middle.sh pride_and_prejudice.txt 20 5`. The script takes the last 5 lines of the first 20 lines, or lines 16 - 20 of pride_and_prejudice.txt. This gives us the title and author of the file, due to the way the text was formatted.

### Q4. More fun with Linux

Try these commands in Bash and interpret the results: `cal`, `cal 2021`, `cal 9 1752` (anything unusual?), `date`, `hostname`, `arch`, `uname -a`, `uptime`, `who am i`, `who`, `w`, `id`, `last | head`, `echo {con,pre}{sent,fer}{s,ed}`, `time sleep 5`, `history | tail`.

**Answer**:

```{bash}
cal
```

Produces a visual calendar of the current UTC date.

```{bash}
cal 2021
```

Produces the full 2021 visual calendar, with the current UTC date annotated.

```{bash}
cal 9 1752
```

Produces an alleged calendar of September 1752. However, this calendar seems to be missing a number of days, namely from September 3, 1752 to September 13, 1752. This happened with Britain changed over from the Julian calendar to the Gregorian calendar in 1752.

```{bash}
date
```

Produces the current UTC date.

```{bash}
hostname
```

The name of the server host. In the case of the server, the name is "ucla-biostat-203b".

```{bash}
arch
```

The machine architecture. In the case of the server, it is "x86_64".

```{bash}
uname -a
```

The information of the computer system, containing OS, server name, date of setup, architecture, etc. The output for the teaching server is "Linux ucla-biostat-203b 3.10.0-1160.6.1.el7.x86_64 #1 SMP Tue Nov 17 13:59:11 UTC 2020 x86_64 x86_64 x86_64 GNU/Linux".

```{bash}
uptime
```

The amount of time the server has been active, along with users simultaneously logged in and the load average for the past minute, 5 minutes, and 15 minutes respectively. The output I got from the server around 1/14/2020 at 5:43 pm is "01:43:32 up 11 days,  3:54,  6 users,  load average: 0.18, 0.16, 0.22".

```{bash}
who am i
```

The user accessing the server and the time logged in. My output is "hunanzhou pts/38       2021-01-15 00:52 (47.145.211.101)".

```{bash}
who
```

The list of all current users logged into the server, including the user's. The output will not be duplicated to preserve the privacy of the other individuals.

```{bash}
w
```

The tabulated and shorthand information from "who" command. Includes information such as IP Address, login time, idle time, etc.

```{bash}
id
```

Information about the current user (in my case, hunanzhou). Includes User ID (uid), group ID (gid), and context.

```{bash}
last | head
```

The top 10 (due to 'head') last logged in users. The output will not be duplicated to preserve the privacy of the other individuals.


```{bash}
echo {con,pre}{sent,fer}{s,ed}
```

Prints out all permutations of words with one part each from each bracketed list. The output comes out to: "consents consented confers confered presents presented prefers prefered". (Note that limitations of the command have resulted in "preferred" being misspelled.)

```{bash}
time sleep 5
```

Pause for 5 seconds and then time the result. Uncertainty in measurement tells us that the result of time may not be exactly 5 seconds. Example output follows:

real	0m5.003s

user	0m0.002s

sys	0m0.001s

```{bash}
history | tail
```

The last 10 (due to 'tail') commands logged in the shell for the given user. For my case, the commands follow:

    372  vi hw1sol.Rmd
    373  last | head
    374  man last
    375  vi hw1sol.Rmd
    376  man last
    377  last | head
    378  vi hw1sol.Rmd
    379  echo {con,pre}{sent,fer}{s,ed}
    380  vi hw1sol.Rmd
    381  history | tail

